---
title: "ToothGrow Analysis"
author: "Enelen Brinshaw"
output:
  pdf_document: 
    fig_height: 3.75
    toc: yes
fontsize : 12pt
---
\newpage

## Overview
This report analyses the effect of different doses and methods of delivery of vitamin C in guinea pigs' odontoblasts (cells responsible for tooth growth). There are three variables in our dataset, `len`, the Tooth length, `supp`, the supplement type, which can be either orange juice (OJ) or ascorbidc acid (VC), and `dose`, the dose in milligrams/day. There are 60 observations, ten for each combination of `supp` and `dose`.    

```{r settings,echo = FALSE,message=FALSE,warning=FALSE}
require(knitr)
knitr::opts_chunk$set(tidy = TRUE,message = FALSE,warning = FALSE)
```

```{r functions,echo = FALSE,tidy = FALSE}
# labeller function to label facets
doses <- function(variable, value){ 
     paste("Dose of", value, "mg/day")
}

# function to print t.test in pander with confidence interval
htest <- function(res, cap = NULL){
     conf <- res$conf.int # extract confidence interval
     attributes(conf) <- NULL # remove its attributes
     conf <- as.character(round(conf, 3)) # and convert to string
     # create a data frame for pander
     temp <- data.frame(df = res$parameter,
                        `Test Statistic` = res$statistic,
                        `p-Value` = res$p.value,
                        `Confidence Interval` = paste(conf[1],
                                                      "$\\quad$",
                                                      conf[2]),
                        check.names = FALSE,
                        row.names = NULL)
     if(is.null(cap)) # if no caption is given use command given
          cap <- paste0("`", res$data.name, "`")
     else 
          cap <- paste("(OJ$-$VC)", cap)
     pander(temp, caption = paste0(res$method, ": ", cap))
}
```


## Exploratory Analysis
Let us first explore the dataset to see the general trends.  
```{r basic,tidy = FALSE}
data("ToothGrowth")
df <- ToothGrowth
df$dose <- as.factor(df$dose)
require(ggplot2)
ggplot(ToothGrowth,aes(dose,len,colour = supp)) +
     geom_point() + geom_smooth() +
     scale_colour_discrete(name = "Delivery Method", 
                           labels = c("Orange Juice (OJ)", "Ascorbic Acid (VC)")) + 
     labs(x = "Dose (mg/day)", y = "Length", 
          title = "Response in Length of guinea pigs to Vitamin C")
```

There is a clear advantage of giving the dose via orange juice rather than ascorbic acid, although the advantage is not present for the highest dose of 2mg/day. Let us look at a few more graphs to gain more understanding.  
```{r boxplot1,tidy = FALSE}
ggplot(df, aes(supp, len, group = supp, fill = supp)) +
     geom_boxplot() +
     facet_wrap(facets = ~dose, labeller = doses) + # user-defined function
     scale_fill_discrete(name = "Delivery Method", 
                         labels = c("Orange Juice (OJ)", "Ascorbic Acid (VC)")) + 
     labs(y = "Length", x = "", 
          title = "Response in length by delivery method for different doses")
```

Again we can see Orange juice seems to be better than Ascorbic acid except at the highest dose.  

## Statistical tests of significance

Let us now confirm whether these differences are significant or due to luck by conducting t-tests, and developing confidence interval for the differences.   
```{r initialTest}
res <- t.test(df$len[df$supp == "OJ"],df$len[df$supp == "VC"])
require(pander)
htest(res,"") # another user defined function see appendix
```
    

The test fails at 95%, but considering there was no difference at the highest dose, this is well within the realms of possibility. Let us redo the tests for each dose.      
   

```{r tests,eval = FALSE,tidy = FALSE}
k <- split(df,f=df$dose)
tests <- lapply(k, function(x) t.test(x$len[x$supp == "OJ"],
                                     x$len[x$supp=="VC"]))
htest(tests[[1]], "Dose = 0.5")
htest(tests[[2]], "Dose = 1.0")
htest(tests[[3]], "Dose = 2.0")
```
```{r ref.label="tests",echo = FALSE}
```

The t-tests confirm our initial exploration. There is a significant difference between tooth length when dose is supplied by Orange juice versus ascorbic acid, but not at the dose level of 2 mg/day (p-value = 0.96 ; interval contains 0). At the level of 0.5mg/day, we are 95% sure, that the orange juice on average produces a greater response in length by an amount between 1.72 to 8.78 when compared to doses given by ascorbic acid. The confidence interval for difference in length for dose of 1 mg/day ranges from 2.8 to 9.1.   

Of course, these are all under the assumptions that the data follows a t-distribution, the samples are not paired, and the variance of the two groups in not equal.   

\newpage 

## Appendix

Code for user-defined functions used in the assignment.   
```{r ref.label="functions",eval = FALSE,tidy=FALSE}
```

